---
title: "day3"
output: html_document
---

--- Spiral Memory ---

You come across an experimental new kind of memory stored on an infinite two-dimensional grid.

Each square on the grid is allocated in a spiral pattern starting at a location marked 1 and then counting up while spiraling outward. For example, the first few squares are allocated like this:

17  16  15  14  13
18   5   4   3  12
19   6   1   2  11
20   7   8   9  10
21  22  23---> ...

While this is very space-efficient (no squares are skipped), requested data must be carried back to square 1 (the location of the only access port for this memory system) by programs that can only move up, down, left, or right. They always take the shortest path: the Manhattan Distance between the location of the data and square 1.

For example:

Data from square 1 is carried 0 steps, since it's at the access port.
Data from square 12 is carried 3 steps, such as: down, left, left.
Data from square 23 is carried only 2 steps: up twice.
Data from square 1024 must be carried 31 steps.
How many steps are required to carry the data from the square identified in your puzzle input all the way to the access port?

Your puzzle input is 265149.

```{r, message=FALSE}
library(dplyr)
board <- data.frame(number = 1:23, 
                    x = c(0,1,1,0,-1,-1,-1, 0, 1, 2,2,2,2,1,0,-1,-2,-2,-2,-2,-2,-1, 0), 
                    y = c(0,0,1,1, 1, 0,-1,-1,-1,-1,0,1,2,2,2, 2, 2, 1, 0,-1,-2,-2,-2))
board
```

```{r}
distance <- function(x, y) {
  sum(abs(x), abs(y))
}

distance(board$x[1], board$y[1]) == 0
distance(board$x[12], board$y[12]) == 3
distance(board$x[23], board$y[23]) == 2
```

This is a problem of odd squares, e.g. `1**2 = 1, 3**2 = 9, 5**2 = 25, 7** = 49,` etc.

```{r}
input <- 265149
next_odd_square <- function(x) {
  y = ceiling(sqrt(x))
  if (y %% 2 == 0) {
    return(y + 1)
  } else {
    return(y)
  }
}
prev_odd_square <- function(x) {
  y = floor(sqrt(x))
  if (y %% 2 == 0) {
    return(y - 1)
  } else {
    return(y)
  }
}
next_odd_square(input)
prev_odd_square(input)
odd_squares <- data.frame(number = seq(1, next_odd_square(input), by = 2)) %>% 
  mutate(square = number**2) %>% 
  mutate(spiral = row_number()) %>% 
  mutate(min_distance = spiral - 1) %>% 
  mutate(max_distance = number - 1)
```


This distance has to be less that the distance of the next square.

For example, at position 12, we know that the square root of 12 is `r sqrt(12)`, 
so the next odd square would be `r next_odd_square(12)` 
and thus the distance has to be less than 
`r odd_squares[odd_squares$number == next_odd_square(12), "distance"]`.

We also know that the minimum distance from the position would be a straight line from the side. 

```{r}
find_distance_range <- function(x) {
  next_sq = next_odd_square(x)
  filter(odd_squares, number == next_sq)
}
find_distance_range(12) # know this is 3
find_distance_range(23) # know this is 2
find_distance_range(49) # know this is 6
find_distance_range(1024) # know this is 31
find_distance_range(input)
```

In order to find the exact distance, we'll need to determine which side of the square the number is on. We know that each additional complete spiral adds the difference between the previous odd square and the next odd square. For example, after 1, the next square (9) adds `9 - 1 = 8` new positions. After that, the next odd square 25 adds `25 - 9 = 14` new positions. 

Since we know the next odd square for our given input, we know how many sides that spiral has at completion. Since we know the previous odd square, we know where to start adding positions from. This helps us thus figure out what side we're on. If the side == 0, then it would be a perfect odd square and thus be the bottom-right corner. If side == 1, then top-right corner, and so on. More so, it also tells us which quadrant we're in and thus the distance to the corner of the square. For example, a side of 3.8 would be the bottom-right quadrant and so we can calculate how far from the corner we are by using that fraction -- e.g. 4 - 3.8 = 0.2, which can be multiplied by the min_distinance. Since we know how far we are from the corner, we know how far we are from the center.

```{r}
distance_info <- function(x) {
  dist_range = find_distance_range(x) %>% 
    mutate(start_position = prev_odd_square(x)**2,
           add_positions = x - start_position,
           side = add_positions/max_distance,
           distance_to_corner = abs(side - round(side, 0)),
           positions_to_corner = max_distance * distance_to_corner,
           positions_to_center = min_distance - positions_to_corner
          )
  return(dist_range)
}
find_distance <- function(x) {
  info = distance_info(x)
  return(info$min_distance + info$positions_to_center)
}
distance_info(input)
find_distance(input)
```

That's the right answer! You are one gold star closer to debugging the printer. 

--- Part Two ---

As a stress test on the system, the programs here clear the grid and then store the value 1 in square 1. Then, in the same allocation order as shown above, they store the sum of the values in all adjacent squares, including diagonals.

So, the first few squares' values are chosen as follows:

Square 1 starts with the value 1.
Square 2 has only one adjacent filled square (with value 1), so it also stores 1.
Square 3 has both of the above squares as neighbors and stores the sum of their values, 2.
Square 4 has all three of the aforementioned squares as neighbors and stores the sum of their values, 4.
Square 5 only has the first and fourth squares as neighbors, so it gets the value 5.
Once a square is written, its value does not change. Therefore, the first few squares would receive the following values:

147  142  133  122   59
304    5    4    2   57
330   10    1    1   54
351   11   23   25   26
362  747  806--->   ...
What is the first value written that is larger than your puzzle input?

Your puzzle input is still 265149.

```{r, message=FALSE}
library(data.table)
board2 <- data.table(
  x = c(0,1,1,0,-1,-1,-1, 0, 1, 2,2,2,2,1,0,-1,-2,-2,-2,-2,-2,-1, 0,1,2), 
  y = c(0,0,1,1, 1, 0,-1,-1,-1,-1,0,1,2,2,2, 2, 2, 1, 0,-1,-2,-2,-2,-2,-2),
  z = c(1,1,2,4,5,10,11,23,25,26,54,57,59,122,133,142,147,304,330,351,362,747,806,NA,NA)
)
```

# add spiral

```{r}
add_spiral_rows <- function(board) {
  board_len <- nrow(board)
  spiral <- next_odd_square(board_len + 1)
  side_len <- max(board[, 1:2], na.rm = TRUE)
  right <- data.frame(
    x = rep(side_len+1, spiral-1),
    y = (-side_len):(side_len+1),
    z = NA)
  top <- data.frame(
    x = (side_len):(-side_len-1),
    y = rep(side_len+1, spiral-1),
    z = NA)
  left <- data.frame(
    x = rep(-side_len-1, spiral-1),
    y = (side_len):(-side_len-1),
    z = NA)
  bottom <- data.frame(
    x = (-side_len):(side_len+1),
    y = rep(-side_len-1, spiral-1),
    z = NA)
  return(bind_rows(board, right, top, left, bottom))
}

add_spiral_value <- function(board) {
  coord <- board[is.na(z)][1]
  near_x <- (coord$x - 1):(coord$x + 1)
  near_y <- (coord$y - 1):(coord$y + 1)
  neighbors <- merge(near_x, near_y, all = TRUE)
  match_neighbors <- merge(neighbors, board)
  coord$z <- sum(match_neighbors$z, na.rm = TRUE)
  # print(coord)
  board[x == coord$x & y == coord$y] <- coord
  return(board)
}
```

```{r}
current_max <- max(board2$z, na.rm = TRUE)
while (current_max < input) {
  board2 <- add_spiral_rows(board2)
  miss_z <- sum(is.na(board2$z))
  while (miss_z > 0) {
    board2 <- add_spiral_value(board2)
    miss_z <- sum(is.na(board2$z))
  }
  current_max <- max(board2$z, na.rm = TRUE)
}
board2[z > input][1]
```

That's the right answer! You are one gold star closer to debugging the printer.
