---
title: "day9"
output: html_document
---

--- Stream Processing ---

A large stream blocks your path. According to the locals, it's not safe to cross the stream at the moment because it's full of garbage. You look down at the stream; rather than water, you discover that it's a stream of characters.

You sit for a while and record part of the stream (your puzzle input). The characters represent groups - sequences that begin with { and end with }. Within a group, there are zero or more other things, separated by commas: either another group or garbage. Since groups can contain other groups, a } only closes the most-recently-opened unclosed group - that is, they are nestable. Your puzzle input represents a single, large group which itself contains many smaller ones.

Sometimes, instead of a group, you will find garbage. Garbage begins with < and ends with >. Between those angle brackets, almost any character can appear, including { and }. Within garbage, < has no special meaning.

In a futile attempt to clean up the garbage, some program has canceled some of the characters within it using !: inside garbage, any character that comes after ! should be ignored, including <, >, and even another !.

You don't see any characters that deviate from these rules. Outside garbage, you only find well-formed groups, and garbage always terminates according to the rules above.

Here are some self-contained pieces of garbage:

```
<>, empty garbage.
<random characters>, garbage containing random characters.
<<<<>, because the extra < are ignored.
<{!>}>, because the first > is canceled.
<!!>, because the second ! is canceled, allowing the > to terminate the garbage.
<!!!>>, because the second ! and the first > are canceled.
<{o"i!a,<{i<a>, which ends at the first >.
```

Here are some examples of whole streams and the number of groups they contain:

```
{}, 1 group.
{{{}}}, 3 groups.
{{},{}}, also 3 groups.
{{{},{},{{}}}}, 6 groups.
{<{},{},{{}}>}, 1 group (which itself contains garbage).
{<a>,<a>,<a>,<a>}, 1 group.
{{<a>},{<a>},{<a>},{<a>}}, 5 groups.
{{<!>},{<!>},{<!>},{<a>}}, 2 groups (since all but the last > are canceled).
```

Your goal is to find the total score for all groups in your input. Each group is assigned a score which is one more than the score of the group that immediately contains it. (The outermost group gets a score of 1.)

```
{}, score of 1.
{{{}}}, score of 1 + 2 + 3 = 6.
{{},{}}, score of 1 + 2 + 2 = 5.
{{{},{},{{}}}}, score of 1 + 2 + 3 + 3 + 3 + 4 = 16.
{<a>,<a>,<a>,<a>}, score of 1.
{{<ab>},{<ab>},{<ab>},{<ab>}}, score of 1 + 2 + 2 + 2 + 2 = 9.
{{<!!>},{<!!>},{<!!>},{<!!>}}, score of 1 + 2 + 2 + 2 + 2 = 9.
{{<a!>},{<a!>},{<a!>},{<ab>}}, score of 1 + 2 = 3.
```

What is the total score for all groups in your input?

```{r, message=FALSE}
library(readr)
library(dplyr)
text = read_lines("day9.txt")
text_chars = unlist(strsplit(text, split = '', fixed = TRUE))
```

```{r}
stream <- data.frame(char = text_chars, stringsAsFactors = FALSE)
```

Remove Cancel (!)

```{r}
cancel_df <- stream %>% 
  mutate(origin = 1:nrow(.)) %>% 
  mutate(exclaim = char == "!")
num_exclaims <- sum(cancel_df$exclaim)
while (num_exclaims > 0) {
  cancel_df <- cancel_df %>%
    mutate(rownum = 1:nrow(.)) %>% 
    mutate(cumsum_exclaim = cumsum(exclaim))
  cancel_rows <- filter(cancel_df, cumsum_exclaim == 1)
  drop_rows <- c(cancel_rows$rownum[1], cancel_rows$rownum[1] + 1)
  cancel_df <- cancel_df[-drop_rows, ]
  num_exclaims <- sum(cancel_df$exclaim)
}
```

Remove Garbage (<>)

```{r}
garbage_df <- cancel_df %>% 
  mutate(start_garbage = char == "<",
         end_garbage = char == ">")
num_garbage <- sum(garbage_df$start_garbage)
while (num_garbage > 0) {
  garbage_df <- garbage_df %>%
    mutate(rownum = 1:nrow(.)) %>% 
    mutate(cumsum_start_garbage = cumsum(start_garbage),
           cumsum_end_garbage = cumsum(end_garbage))
  garbage_rows <- filter(garbage_df, cumsum_start_garbage >= 1 & cumsum_end_garbage < 1)
  drop_rows <- c(garbage_rows$rownum, max(garbage_rows$rownum) + 1)
  garbage_df <- garbage_df[-drop_rows, ]
  num_garbage <- sum(garbage_df$start_garbage)
}
```

Score Groups ({})

```{r}
score_df <- garbage_df %>% 
  filter(char == "{" | char == "}") %>% 
  select(char) %>% 
  mutate(open = char == "{",
         close = char == "}")
score <- 0
num_open <- sum(score_df$open)
while (num_open > 0) {
  score_df <- score_df %>% 
    mutate(rownum = 1:nrow(.)) %>% 
    mutate(score = cumsum(open))
  first_close <- filter(score_df, close)[1,]
  score_group <- filter(score_df, rownum <= first_close$rownum)
  group_rows <- filter(score_df, score == max(score_group$score))[1:2,]
  drop_rows <- group_rows$rownum
  score <- score + group_rows$score[1]
  score_df <- score_df[-drop_rows, ]
  num_open <- sum(score_df$open)
}
```

That's the right answer! You are one gold star closer to debugging the printer.

--- Part Two ---

Now, you're ready to remove the garbage.

To prove you've removed it, you need to count all of the characters within the garbage. The leading and trailing < and > don't count, nor do any canceled characters or the ! doing the canceling.

```
<>, 0 characters.
<random characters>, 17 characters.
<<<<>, 3 characters.
<{!>}>, 2 characters.
<!!>, 0 characters.
<!!!>>, 0 characters.
<{o"i!a,<{i<a>, 10 characters.
```

How many non-canceled characters are within the garbage in your puzzle input?

```{r}
garbage_rows <- cancel_df %>% 
  mutate(in_garbage = !(origin %in% garbage_df$origin)) %>% 
  filter(in_garbage) %>% 
  mutate(open = char == "<",
         close = char == ">")
num_garbage <- sum(garbage_rows$open)
count_garbage <- 0
while (num_garbage > 0) {
  garbage_rows <- garbage_rows %>%
    mutate(rownum = 1:nrow(.)) %>% 
    mutate(cumsum_open = cumsum(open),
           cumsum_close = cumsum(close))
  count_rows <- filter(garbage_rows, cumsum_open >= 1 & cumsum_close < 1)
  count_garbage <- count_garbage + nrow(count_rows) - 1
  drop_rows <- c(count_rows$rownum, max(count_rows$rownum) + 1)
  garbage_rows <- garbage_rows[-drop_rows, ]
  num_garbage <- sum(garbage_rows$open)
}
```

That's the right answer! You are one gold star closer to debugging the printer.
